```{r setup, include=FALSE, echo=FALSE, message=FALSE}
# set global chunk options
opts_chunk$set(cache=FALSE, message=FALSE)
options(stringsAsFactors=FALSE)
```

```{r data_setup, echo=FALSE}
library(rbenchmark)
library(Hmisc)
library(HBA)
library(data.table)
PermuteCol <- function(x) {
   attrx <- attributes(x)
   x <- x[sample(1:length(x), length(x)), drop=FALSE]
   attributes(x) <- attrx  
   x
}

set.seed(1005)
ptdemog[] <- lapply(ptdemog, PermuteCol)
ptdemog <- subset(ptdemog, , c("RID", "SITEID", "PTAGE", "PTGENDER", "PTEDUCAT"))
mmse[] <- lapply(mmse, PermuteCol)
mmse <- subset(mmse, , c('RID', "VISCODE", "MMSCORE"))
cogp1[] <- lapply(cogp1, PermuteCol)
cogp1 <- subset(cogp1, , c("RID", "VISCODE", paste0("COT", 1:3, "SCOR"), paste0("TICS", 1:8), "TICTOTAL"))
```
#### Table of Contents
[data.table] (#data.table)
   [data.table basic syntax] (#data.table.syntax)
   [data.table cookbook] (#data.table.cookbook)
[dplyr] (#dplyr)

<a name="data.table"/>
# data.table
## build to reduce 2 types of time
1. programming time (easier to write, read, debug and maintain)
2. compute time

## data.table, just like a data.frame. However :
1. it never has rownames. Instead it may have **one key of one or more columns**. 
This key can be used for row indexing instead of rownames, and **duplicate key values are allowed**.
2. it has enhanced functionality in `[.data.table` for fast joins of keyed tables, 
fast aggregation, fast last observation carried forward (LOCF) and fast 
add/modify/delete of columns by reference with no copy at all.

<a name="data.table.syntax"/>
## Basic syntax
x[i, j, by] - Take `x`, subset rows using `i`, then calculate `j` grouped by `by`
- i: *Integer, logical* or *character vector*, *expression of column names, list* or *data.table*
   - Integer and logical vectors work the same way they do in `[.data.frame`, but
     be cautious with `NA`.
   - character: matched to the first column of x's key
   - expression: evaluated within the frame of the data.table
   - data.table: x must have a key. i is _joined_ to x using x's key and the rows 
     in x that match are returned. The number of join columns is determined by 
     min(length(key(x)), if (haskey(i)) length(key(i)) else ncol(i))
 
- j: A single column name, single expresson of column names, list() of expressions of column names, 
     an expression or function call that evaluates to list is expression or list of expressions, 
     evaluated with data.table. _When with=FALSE a vector of names or positions to select_

- by: A single unquoted column name, a list() of expressions of column names, 
      a single character string containing comma separated column names.

**R: `i, j, by`**
**SQL: `WHERE, SELECT, GROUP BY`**

<a name="data.table.cookbook"/>
## Cookbook 
### Basic Operations
```{r data.table basic, echo=TRUE, eval=FALSE}
DF <- ptdemog 
DT <- data.table(ptdemog)
DT_mmse <- data.table(mmse)
DT_cogp1 <- data.table(cogp1)
head(DF)
DT
identical(dim(DT),dim(DF)) # TRUE
identical(DF$a, DT$a)      # TRUE
is.list(DF)                # TRUE
is.list(DT)                # TRUE
is.data.frame(DT)          # TRUE

tables()                   # see a list of all data.tables in memory 

# We can use data.frame syntax in a data.table, when no keys have been set 

DT[DT$RID == 226, ]               
DT[2, ]                          # 2nd row
DT[2]                            # 2nd row
DT[, 5]                          # not the 5th column, j is expression so 5 is 5..
DT[, 5, with=FALSE]              # the 5th column, j can be number when with=FALSE, default is TRUE.
colNum = 5
DT[, colNum, with=FALSE]         # same
DT[, PTAGE]                      # PTAGE column (as vector)
DT[, list(PTAGE)]                # PTAGE column (as data.table)
DT[, "PTAGE", with=FALSE]        # same
DT[, list(RID, VISCODE, PTAGE)]  # choose multiple columns

setkey(DT, RID)                  # DT will be sorted with key
setkeyv(DT, "RID")               # DT will be sorted with key
setkey(DT_cogp1, RID, VISCODE)   # set key with multiple columns, Sorts the table by RID then VISCODE 
setkeyv(DT_cogp1, c("RID", "VISCODE"))
key(DT)
DT[226]                          # row with RID == 226, not 226th row
DT[226, ]                        # same 
DT[NA]                           # not recycle to match the number of rows
DT[NA, ]                         # same 
DF[NA, ]                         # recycle to match the number of rows
DT[nrow(DT) + 1, ]               # index i out of range
DF[nrow(DF) + 1, ]               # same behavior

DT[DT$RID == 226, ]              # vector scan (slow) but same
DT_cogp1[list(226)]              # rows with RID == 226, binary search (fast)
DT_cogp1[J(226)]                 # same 
DT_cogp1["226"]                  # same 
DT_cogp1[226]                    # Be careful, it's not rows with RID == 226
tables()

DT[2:5, mean(PTAGE)]          # sum(PTAGE) over rows 2 and 3
DT[2:5, paste0(RID,"\n")]     # just for j's side effect
DT[c(FALSE,TRUE)]             # even rows (usual recycling)
# COT2SCOR, COT3SCOR, 
DT_cogp1[, mean(COT1SCOR, na.rm=TRUE), by=list(RID, VISCODE)]   # keyed by
DT[ , sum(v), by=key(DT)]     # same
DT[ , sum(v), by=y]           # ad hoc by

DT[, PTAGE_n:=scale(PTAGE, center = TRUE, scale = TRUE)]         #vanilla update note the := operator 
DT[,`:=`(PTAGE_c=mean(PTAGE, na.rm=TRUE), PTAGE_s=PTAGE_n - 1)]  #update several columns at once
DT[, c("USERID", "USERDATE", "USERID2", "USERDATE2"):=NULL]      #remove several columns at once


```

## Merge and Join
- `X[Y, nomatch=NA]`: all rows in Y, *right outer join* (default) X[Y]
  - `merge(X, Y, all.y=TRUE)`
- `X[Y, nomatch=0]`: only rows with matches in both X and Y, *inner join*
  - `merge(X, Y, by=key)`
- `Y[X]`: all rows in X, *left outer join*
  - `merge(X, Y, all.x=TRUE)`
- `unique_keys <- unique(c(X[,t], Y[,t])); Y[X[J(unique_keys)]]`, all rows from both X and Y - full outer join 
- or `X[Y[J(unique_keys)]]` 
  - `merge(X, Y, all=TRUE)`

```{r data.table join, echo=TRUE}
# from data.table examples
DF <- data.frame(x=rep(c("a","b","c"),each=3), y=c(1,3,6), v=1:9)
DT <- data.table(x=rep(c("a","b","c"),each=3), y=c(1,3,6), v=1:9, key='x')
DT
X <- data.table(c("b","c"), foo=c(4,2))
X
DT["a", sum(v)]                  # j for one group
DT[c("a","b"), sum(v)]           # j for two groups

DT[X]                            # join 
DT[X, sum(v)]                    # join and eval j for each row in i
DT[X, mult="first"]              # first row of each group
DT[X, mult="last"]               # last row of each group
DT[X, sum(v)*foo]                # join inherited scope

setkey(DT,x,y)                   # 2-column key
setkeyv(DT, c("x","y"))           # same

DT["a"]                          # join to 1st column of key
DT[J("a")]                       # same. J() stands for Join, an alias for list()
DT[list("a")]                    # same
DT[.("a")]                       # same. In the style of package plyr.
DT[J("a",3)]                     # join to 2 columns
DT[.("a",3)]                     # same
DT[J("a",3:6)]                   # join 4 rows (2 missing)
DT[J("a",3:6), nomatch=0]        # remove missing
DT[J("a",3:6), roll=TRUE]        # rolling join (locf)

DT[, sum(v), by=list(y%%2)]      # by expression
DT[, .SD[2], by=x]               # 2nd row of each group
DT[, tail(.SD, 2), by=x]         # last 2 rows of each group
DT[, lapply(.SD, sum), by=x]     # apply through columns by group

DT[, list(MySum=sum(v),
          MyMin=min(v),
          MyMax=max(v)),
by=list(x,y%%2)]                 # by 2 expressions

DT[, sum(v), x][V1<20]           # compound query
DT[, sum(v), x][order(-V1)]      # ordering results

print(DT[, z:=42L])              # add new column by reference
print(DT[, z:=NULL])             # remove column by reference
print(DT["a", v:=42L])           # subassign to existing v column by reference
print(DT["b", v2:=84L])          # subassign to new column by reference (NA padded)

DT[,m:=mean(v), by=x][]          # add new column by reference by group
                                 # NB: postfix [] is shortcut to print()

DT[,.SD[which.min(v)], by=x][]   # nested query by group, .SD - Subset of Data.table

DT[!J("a")]                      # not join
DT[!"a"]                         # same
DT[!2:4]                         # all rows other than 2:4
DT[x!="b" | y!=3]                # multiple vector scanning approach, slow
DT[!J("b",3)]                    # same result but much faster

```

```{r ADCS task cookbook, echo=TRUE}
DT_mmse = data.table(mmse)
DT_cogp1 = data.table(cogp1)
DT_demog = data.table(ptdemog)
setkey(DT_demog, RID)                 
setkey(DT_cogp1, RID, VISCODE)   # set key with multiple columns, Sorts the table by RID then VISCODE 
setkey(DT_mmse, RID, VISCODE)   # set key with multiple columns, Sorts the table by RID then VISCODE 

# add a new columns treatment
DT_demog[, treatment:= ifelse(sample(c(0, 1), nrow(DT_demog), replace=TRUE), 'treatment', 'placebo')]
COTs <- colnames(DT_cogp1)[grepl('COT\\d+SCOR', colnames(DT_cogp1), perl=TRUE)]
cols <- c('RID', 'VISCODE', COTs)
DT_join <- DT_demog[, c('RID', 'treatment', 'PTAGE'), with=FALSE][DT_cogp1[, cols, with=FALSE]]
DT_merge <- merge(DT_demog[, c('RID', 'treatment', 'PTAGE'), with=FALSE], DT_cogp1[, cols, with=FALSE], by='RID', all.y=TRUE)
identical(DT_join, DT_merge)

DT_join[, lapply(.SD[, COTs, with=FALSE], function(x) mean(x, na.rm=TRUE)), by='treatment'] # not preferred 
DT_join[, lapply(.SD, function(x) mean(x, na.rm=TRUE)), by='treatment', .SDcols=COTs]       # preferred
DT_join[, lapply(.SD, function(x) mean(x, na.rm=TRUE)), by='VISCODE', .SDcols=COTs]

# mean of each COTs by VISCODE and treatment
head(DT_join[, lapply(.SD, function(x) mean(x, na.rm=TRUE)), by='VISCODE,treatment', .SDcols=COTs])
# total counts by VISCODE and treatment 
head(DT_join[, lapply(.SD, function(x) length(x)), by='VISCODE,treatment', .SDcols=COTs])

# non missing total counts by VISCODE and treatment 
head(DT_join[, lapply(.SD, function(x) {sum(! is.na(x))}), by='VISCODE,treatment', .SDcols=COTs])
# the output is differnt
head(DT_join[, list(lapply(.SD, function(x) {sum(! is.na(x))})), by='VISCODE,treatment', .SDcols=COTs])

# calculate both missing and non missing total counts by VISCODE and treatment
# with verbose=TRUE
DT_join[, lapply(.SD, function(x) {
                 list(sum(! is.na(x)), sum(is.na(x)))
          }), by='VISCODE,treatment', .SDcols=COTs]
DT_join[, list(lapply(.SD, function(x) {sum(! is.na(x))}), 
               lapply(.SD, function(x) {sum(is.na(x))})), by='VISCODE,treatment', .SDcols=COTs, verbose=TRUE]

```

## Reference
- [data.table R forge] (http://datatable.r-forge.r-project.org/)
- [data.table Reference manual] (http://cran.r-project.org/web/packages/data.table/data.table.pdf)
- [stackoverflow for data.table] (http://stackoverflow.com/questions/tagged/data.table)
- [data.table wiki] (http://rwiki.sciviews.org/doku.php?id=packages:cran:data.table)

# dplyr

## Basic syntax
 - use dplyr to transform data
   * filter: return rows that meet some criteria
   * select: return subset of columns
     + starts_with(x): names starts with "x"
     + ends_with(x): names ends with "x"
     + contains(x): select all variables whose name contains "x"
     + matches(x): select all variables whose names matches regular expression "x"
     + num_range("x", 1:5, width=2): select all variables from x01 to x05.
     + use `-` to drop variables.
   * arrage: reorder rows
   * mutate: add new columns
   * summarise: reduce each group to a single row.
 - first argument is data.frame and always return data.frame (no drop).

```{r dplyr demo data, echo=FALSE, eval=TRUE}
   library(dplyr)
   library(ggplot2)
   flights <- tbl_df(read.csv("flights.csv", stringsAsFactors = FALSE))
   flights$date <- as.Date(flights$date)
   weather <- tbl_df(read.csv("weather.csv", stringsAsFactors = FALSE))
   weather$date <- as.Date(weather$date)
   planes <- tbl_df(read.csv("planes.csv", stringsAsFactors = FALSE))
   airports <- tbl_df(read.csv("airports.csv", stringsAsFactors = FALSE))
```

### Basic Operations
```{r dplyr basic, echo=TRUE, eval=FALSE}
filter(flights, dest %in% c("SFO", "OAK")) 
filter(flights, dest == "SFO" | dest == "OAK")
by_color <- group_by(df, color)
summarise(by_color, total = sum(value))
```

## Reference
- [useR2014 Hadley's dplyr tutorial] (https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a)
